---
title: "Statistical Inference"
author: "Stephen Wade"
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
output:
  pdf_document:
    includes:
      in_header: CP1.sty
  html_document: default
fontsize: 12pt
---

# Effect of Vitamin C on Tooth Growth in Guinea Pigs

## Outline
A study was conducted on sixty guinea pigs measuring the length of cells
responsible for tooth growth, called odontoblasts, given a certain treatment of
vitamin C. The guinea pigs received one
of three dose levels (0.5, 1 and 2 mg/day) by one of two delivery methods,
orange juice or ascorbic acid. Thus there are six groups of ten guinea pigs, 
where each group received a particular dosage/delivery method.

The three variables in the dataset are `len`, the tooth length (units not
specified), `supp`, which is the delivery method, and `dose` which is the dose
in mg/day.

The dataset is supplied by the `datasets` package in R, and the summary
above is found [here](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html).

The question we wish to answer is does delivery method or dosage
affect the odontoblast measurement?

## Exploratory Analysis

A basic exploratory analysis is to consider a boxplot and a summary of the means
and standard deviations of each group.

```{r exploratory, fig.width=7, fig.height=3, fig.align="center", echo=FALSE}
 library(ggplot2); library(dplyr); library(knitr); library(datasets)

g <- ggplot(ToothGrowth,
            aes(x=factor(dose),y=len),
            title='Exploratory plot of tooth growth dataset')
g + geom_boxplot() + facet_grid(.~supp) +
    xlab('dose mg/day') + ylab('length') + theme_bw()
```

```{r summary_table, echo=FALSE}
kable(summarize(group_by(ToothGrowth, supp, dose),
                mean_length=mean(len), sd_length=sd(len), count=n()))
```

From the boxplot it can be seen that increasing dosage does
seem to increase the odontoblast length measurement. The medians of the length of
each group increases as the dosage increases in each panel.

The summary of the means and the standard deviations shows increasing dosage
also implies increasing length, and the standard deviation does not appear
correlated with either dosage or delivery method.

It is unclear as to whether the effect of the delivery is significant from this
exploratory analysis, except as clearly seen at the dosage equal to 1mg/day
level by comparing panels.

## Hypothesis tests
We will test the hypothesis that the dosage does not affect growth of
odontoblasts given a method of delivery. For this we assume the measurements are i.i.d. distributed within
groups with potentially unequal variances. The appropriate test is then a two-sided t-test with `unequal=TRUE` (see code in
the [Appendix](#appendix)).

Similarly, we will test the hypothesis that the method of delivery at fixed
dosages does not have an effect on odontoblast length using the same
assumptions.

Given the small number of tests and intuitive treatments, no correction 
(e.g. Bonferroni or other p-value adjustment) is
considered for family-wise-error-rate (FWER) or false discovery rate.

```{r tests,echo=FALSE}
pvs <- pairwise.t.test(ToothGrowth$len,
                       interaction(ToothGrowth$dose,ToothGrowth$supp),
                       pool.sd=FALSE,
                       p.adjust="none")$p.value
OJpvs <- pvs[c('1.OJ','2.OJ'),c('0.5.OJ','1.OJ')]
rownames(OJpvs) <- c('1mg/day','2mg.day')
colnames(OJpvs) <- c('0.5mg/day', '1mg/day')
VCpvs <- pvs[c('1.VC','2.VC'),c('0.5.VC','1.VC')]
rownames(VCpvs) <- c('1mg/day','2mg.day')
colnames(VCpvs) <- c('0.5mg/day', '1mg/day')
DOSEpvs <- t(diag(pvs[c('0.5.VC','1.VC','2.VC'),
                      c('0.5.OJ','1.OJ','2.OJ')]))
colnames(DOSEpvs) <- c('0.5mg/day', '1mg/day','2mg/day')
```

### Dosage effect given delivery via orange juice

The p-values for the difference in mean between dosage groups of guinea pigs
given delivery by orange juice is given below. The main result is that all
three values are significant. 
There may be a decreasing effect of odontoblast growth as dosage
increases (no regression was performed to justify this, this is merely an
observation), as observed by the larger p-value for the comparison of
1mg/day compared to 2mg/day.

```{r OJtable,echo=FALSE}
kable(OJpvs,
      caption=paste('P-values $\\mu_{\\text{doseA}} - \\mu_{\\text{doseB}}$',
                    'given Orange juice delivery'))
```

### Dosage effect given delivery via ascorbic acid

The p-values for the difference in mean between dosage groups of guinea pigs
given delivery by ascorbic acid is given below. The main result is that all
three values are significant, with very low probabilities of observing
as or more extreme statistics given the hypothesis that the groups are the
same.

```{r VCtable,echo=FALSE}
kable(VCpvs,
      caption=paste('P-values $\\mu_{\\text{doseA}} - \\mu_{\\text{doseB}}$',
                    'given ascorbic acid delivery'))
```

### Delivery effect by dosage

The p-values for the difference in mean between delivery given a fixed dosage
is shown below. Here the results indicate that at lower dosage, the
delivery mechanism is an important factor, however a p-value of
`r DOSEpvs[3]` for the 2mg/day dosage suggests that at this level of dosage,
there is a high probability of observing the statistic obtained under
the hypothesis that the groups are the same.

```{r DOSEtable,echo=FALSE}
kable(DOSEpvs,
      caption=paste('P-values $\\mu_{\\text{OJ}} - \\mu_{\\text{VC}}$',
                    'by dosage level'))
```

## Conclusions and further remarks

The results suggest that the dosage level has an effect on odontoblast length
given either method of delivery. The effect of delivery is significant at
the two lower dosage levels, 0.5mg/day and 1mg/day, whilst at the highest dosage
level there was no significant evidence to suggest any difference in
odontoblast length due to the method of delivery.

\pagebreak

# Appendix

The code to produce this report is given below.

```{r ref.label='exploratory',eval=FALSE}
```
```{r ref.label='summary_table',eval=FALSE}
```
```{r ref.label='tests',eval=FALSE}
```
```{r ref.label='OJtable',eval=FALSE}
```
```{r ref.label='VCtable',eval=FALSE}
```
```{r ref.label='DOSEtable',eval=FALSE}
```