---
title: "Statistical Inference"
author: "Stephen Wade"
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
output:
  pdf_document:
    includes:
      in_header: CP1.sty
  html_document: default
fontsize: 12pt
---

# Exponential RV's and the Central Limit Theorem

## Outline
An exponential random variable is a model for the waiting
time of a Poisson process with rate $\lambda$. Using the Heaviside 'step' function $\text{H}(x)$, the probability distribution 
(that is, the cumulative distribution function) of an exponential variable is
$$ F(x;\lambda) = (1-e^{-\lambda x}) \text{H}(x) $$

Consider $n$ independently identically distributed (i.i.d.) exponential
random variables with rate $\lambda$, which we denote $X_i$. Observations made
of these random variables are given in lower case $x_i$. Let $\bar{X}$
represent the mean,
$$ \bar{X} = \sum{x_i}{n}, $$
which is also a random variable. It is an estimator of the expected value of the exponential distribution $1/\lambda$ because
$$ \mu_{\bar{X}} = \text{E}\left[\bar{X}\right] = \frac{1}{\lambda}. $$
Futhermore, the expected value of the standard error $\sigma_{\bar{X}}$ is found by
$$ {\sigma_{\bar{X}}}^2 =
    \text{E}\left[ \left(\bar{X} - \mu_{\bar{X}}\right)^2\right] =
    \frac{1}{n\lambda^2}. $$

### Central limit theorem
The central limit theorem (CLT) states that as $n$ gets large
  $$ \frac{\bar{X} - \mu_{\bar{X}}}{\sigma_{\bar{X}}} \sim N(0,1) $$
  
### Aim and methodology
The aim of this exercise is to provide evidence of the central limit theorem.
A sample mean is calculated by using $n=40$ pseudo-randomly generated
observations of the exponential distribution with $\lambda=0.2$. This process 
is repeated $1000$
times so we then have a pseudo-randomly generated sample distribution of
$\bar{X}$, that is the distribution of averages of 40 exponential distributions
with $\lambda=0.2$. Code is given in 
the [Appendix](#appendix).

## Results

### Sample mean and variance 
We wish to check that
the sample mean of our sample from the distribution of $\bar{X}$ is, as per the
central limit theorem, $1/\lambda=5$. We also wish to check that the sample
variance of our sample from the distribution of $\bar{X}$ is
$1/n\lambda^2=25/40$. The theoretical versus sample results are shown in the
table below.

```{r generate_means, echo=FALSE}
library(knitr)
set.seed(1234321)
nosim <- 1000
n <- 40
lambda <- 0.2
sim_exp <-matrix(rexp(nosim * n,rate=lambda), nosim)
sxbar <- apply(sim_exp, 1, mean)
```

```{r generate_table, echo=FALSE}
sample_m <- mean(sxbar)
theoretical_m <- 1 / (lambda)
sample_var <- var(sxbar)
theoretical_var <- 1 / (n * lambda^2)
error_m <- abs(sample_m-theoretical_m)/theoretical_m
error_var <- abs(sample_var-theoretical_var)/theoretical_var

kable(data.frame(theoretical=c(theoretical_m,theoretical_var),
           sample=c(sample_m, sample_var),
           'rel error'=c(error_m, error_var),
           row.names=c('mean$(\\bar{X})$','var$(\\bar{X})$')))
```

As the table shows, the relative error is very small for both values, and
hence the sample mean and sample variance of the sample from the distribution
of $\bar{X}$ (the averages of 40 exponentials with $\lambda=0.2$) is very
close to that expected by the CLT.


### Distribution of $\bar{X}$

We now wish to show that the distribution of $\bar{X}$ is approximately 
normal as per the CLT. The histogram on the next page shows the distribution of our sample
of $$ \frac{\bar{X} - \mu_{\bar{X}}}{\sigma_{\bar{X}}} $$ with a blue line
showing the ideal $N(0,1)$ standard normal distribution. The $y$ axis is
the density (rather than count), and the comparison is very reasonable, 
suggesting that the distribution is close to normal.

\pagebreak

```{r histogram_CLT, fig.width=5, fig.height=4, fig.align="center", echo=FALSE}
library(ggplot2)
sxbnorm <- data.frame(means=(sxbar - 1/lambda) / (1/(sqrt(n)*lambda)))
g <- ggplot(data=sxbnorm, aes(means))
g + geom_histogram(aes(y = ..density..),
                   binwidth=0.25,
                   color='light grey') + 
    xlab('normalised sample mean') +
    scale_x_continuous(limits=c(-3,3)) +
    stat_function(fun = dnorm, geom = "line",
                  colour='cornflowerblue',
                  linetype=2, 
                  size=1) +
    theme_bw()
```

As a bonus, I've included a quantile-quantile plot below which shows how the quantiles of the data compare to a normal distribution, the solid line being
where the 
samples (dots) should be if it were exactly normal. As is often the case, some
non-normality is shown at the tails of the distribution, however this would
decrease if $n$ were larger.

```{r qqplot_CLT, fig.width=5, fig.height=4, fig.align="center", echo=FALSE}
library(ggplot2)
g <- ggplot(data=sxbnorm, aes(sample=means))
g + geom_abline(slope=1,
                intercept=0,
                colour='cornflowerblue',
                alpha=0.75) +
    stat_qq(size=1.5, alpha=0.35) + 

    theme_bw()
```

\pagebreak

# Appendix

The code to produce this report is given below.

```{r ref.label='generate_means', eval=FALSE}
```
```{r ref.label='generate_table', eval=FALSE}
```
```{r ref.label='histogram_CLT', eval=FALSE}
```
```{r ref.label='qqplot_CLT', eval=FALSE}
```
